Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl

### 2023-04-06 Code/Draft_ruddy_duck_vital_rates2.Rmd
Second version submitted to coauthors

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]

@@

  **Keywords**: fecundity - productivity - reproductive success - age ratio - duck
output: 
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
    bookdown::word_document2:
      toc: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

We monitored the demography of Ruddy duck populations in both Great-Britain and France thanks to exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition to these exhaustive counts, censuses that distinguished female-like from male-like individuals (Figure \@ref(fig:count)) were performed from 2006 to 2012 in Great Britain, which corresponded to a strong decline of the population. In France, the monitoring spread over 1999, 2001-2009, and 2012-2019. This population grew rapidly during the first years of the monitoring and then stabilized from 2006, which corresponds to a period with a strong control effort. During these years, the eradication effort, under control of the authorities in both countries, was recorded. Even if the age ratio in the hunting bag was unknown before 2006 in France, we can qualify the harvest pressure on adults in three categories for both countries: "no harvest" before 1999, "low harvest" between 1999 and 2005, and "high harvest" from 2006 (see Figure \@ref(fig:sup) in Supplementary Materials).

@@

In waterfowl, the survival is commonly higher in adult males than in adult females. Adult sex ratio is then generally biased towards males (Figure \@ref(fig:pfem), @Wood2021). As a consequence, deducing the immature proportion within a population from the observed proportion of adult males is not straightforward because the adult proportion is not just twice as much as the observed proportion of adult males. There were no counts differentiating male-like from female-like during the breeding period to estimate properly adult sex ratio. But sex identification in adults from hunting bags was available for both Ruddy duck populations. However, the small population size in France prevented from getting precise adult sex ratios. A comparison over months in UK showed no difference in adult male proportion, so data from adults collected over the whole year were used to estimate the male proportion. A comparison of male proportion among years with more than 500 samples did not show significant interannual differences. Even if the adult sex ratio may vary over a long time range, it is relatively stable over a few years [@Wood2021]. As a consequence, hunting bag data from all adults were pooled to estimate the male proportion in adults, see Equation \@ref(eq:pm).

@@

The maximum growth rate occurred for both populations during their expansion phase, when the evolution of the population size was the steepest. To estimate robust maximum population growth rates for both populations, we smoothed annual population growth rate over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L). To do so, we rejected Great Britain data before 1972 since the size estimate of this population is noisy below 50 individuals (Figure \@ref(fig:count)). After reaching the threshold of 1000 individuals, the Great Britain population growth observed a strong inflexion whereas no harvest pressure was applied. This observation suggests that a negative density-dependence process might occur beyond 1000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in Great Britain, i.e. 1972-1981. For the French population, the sequence without harvest spread over the period 1994-1999.

@@

Even if both vital rates vary in a likely range following the validation methodology previously described, it does not prove that the interannual variability is properly tracked. In order to validate this aspect, we compared the outputs of the count-based method, i.e. the immature proportion and both vital rates, to outputs based on data from hunting bags. Only data from years with more than 100 individuals harvested in winter were selected. It covers 5 years of the count time series of the Great Britain population. The presence of the bursa of Fabricius enables immature to be identified [@Hochbaum1942]. The estimation of the immature proportion in hunting bags is then straightforward, see Equation \@ref(eq:pis). The adult survival and the recruitment rate were then estimated by combining this proportion and Equation \@ref(eq:vr). We discussed the ability of the count-based method to accurately track the interannual variability by comparing the outputs of the two methods.

@@

## Impact of harvest strategies on population dynamics

Over the "high harvest" period, i.e. from 2006, control strategies differed from Great Britain to France. In Great Britain, harvest occurred mostly in winter, i.e. before the breeding period (53.4% of adults were shot before the 30^th May), whereas it was done mostly in summer in France, i.e. during and after the breeding period (81.6% of adults were shot after the 30^th May). We compared the population growth of the two populations in light of the difference between the two strategies and we assessed the respective response of the vital rates by comparing the average values over the "high harvest" period to proxies of maximum vital rates estimated when both populations reached their maximum growth (see Section \@ref(VAL)).

For all the sub-models in Section \@ref(MM), the Bayesian framework was used for its efficiency and simplicity to propagate error through the parameters. We used uninformative priors on all parameters. As the maximum growth rate is an life history trait quite stable among populations of a same species, we used a uninformative hierarchical prior for this parameter. We generated three chains of length 500000, with a thinning of 10 to avoid autocorrelation in the samples, and we discarded the first 2000 samples as burn-in. Chain convergence was assessed using the Gelman and Rubin convergence diagnostic (R<1.1, @Gelman1992). We fit the models using NIMBLE [@Valpine2017] run from R [@RCT2022]. Data and code are available here: https://github.com/adri-tab/Ruddy_duck_vital_rates. The values **X[Y; Z]** reported in results are the medians and the associated boundaries of the 95% confidence interval of posterior distributions. The median was preferred to the mean because of its robustness to skewed distribution.

@@

The harvest rate on adults is considered high for both populations from 2006 (Figure \@ref(fig:sup)). There is no significant difference between the two populations, t-test p-value : 0.23, and the average harvest rate is 0.49.

Time series derived from count data show that the two populations display comparable initial growth. They grew freely in both countries until 1999. Although shooting was initiated in 1999 in both countries, both populations begun declining only after much higher harvest efforts had been achieved in 2005 and 2018 respectively (figure 3). This history underlines that both populations observed a large spectrum of harvest effort, from no pressure to very high pressure. For sake of interpretation, we ascribed the time-period 2001-2004 to the no harvest category in France because the harvest pressure was negligible hand hence had no effect on the population growth.

Also, covering contrasted populations dynamics corresponding to different levels of harvesting pressure. In France, data before 2004 correspond to a period of steady population growth with (almost) no harvest pressure, whereas data in the following years correspond to a stabilized population size with a significant level of harvest pressure . In UK, the counting data correspond to a quick population depletion associated to a high level of harvest pressure, especially before the reproduction period. In parallel, the sampling dataset, which corresponds to individuals shot in winter during control operations, covers 9 years only in UK, with 5 years in common with the corresponding counting time series.
Estimating recruitment rate from the evolution of apparent sex-ratio in winter

Time series show that the two populations follow similar demographic histories (Figure \@ref(fig:count)). They grew freely in both countries until 1999. Control measures have been then applied from 1999 to 2005 in the UK and from 1999 to 2018 in France. In both countries, these management measures led to stop the expansion of the ruddy ducks. From 2005 in the UK and from 2018 in France, higher control effort led to a rapid depletion in both populations. For sake of interpretation, the time period 2001-2004 was classified in the *no harvest* category in France because the harvest pressure applied over this period was very limited with no effect on the population growth.

The counting dataset, which differentiates the male-like from the female-like individuals in winter, covers 7 years and 19 years in UK and France respectively. Combined together, the counting datasets cover different population dynamics corresponding to a gradient of harvest pressure. In France, data before 2004 correspond to a period of population growth with (almost) no harvest pressure, whereas data in the following years correspond to a stabilized population size with a significant level of harvest pressure. In UK, the counting data correspond to a quick population depletion associated to a high level of harvest pressure, especially before the reproduction period. In parallel, the sampling dataset, which corresponds to individuals shot in winter during control operations, covers 9 years only in UK, with 5 years in common with the corresponding counting time series.

## Likelihood of the range of vital rates

########## OLD Recruitment rate variability

The male proportion in adults, which is necessary to infer on the recruitment rate from count data, is estimated at 0.60 [0.59; 0.61]. The counting method provides estimations of the recruitment rate, see Figure \@ref(fig:vital). In UK, values vary from 0.06 [0.05; 0.07] to 0.27 [0.20; 0.34] recruits per breeder, and in France, 0.08 [0.04; 0.12] to 0.76 [0.64; 0.88] recruits per breeder. There are significant year-to-year variations with a maximum amplitude of 0.68 recruits per breeder in France (9.5-time variability). On the available period, recruitment rate in UK is smaller than in France. Recruitment rates are estimated in both populations over 5 years, and no correlation is observed between the two populations. The lowest recruitment rate observed in France, which is in 2019, corresponds to the start of the depletion of the population. It is noticeable that it is in the same range than the UK values, also estimated during a strong population depletion.

Recruitment rates estimated from UK sample data are consistently higher than estimates from the corresponding count data, Figure \@ref(fig:vital). The correlation between the two methods is strong and points out that the sampling method overestimates the recruitment rate by 44% compared to the counting method, Figure \@ref(fig:vital). Estimation precision is similar for the two methods. The UK population started to deplete from 2006, but no significant shift is observed between the two periods.

@@

## Likelihood of interannual variability in vital rates

########## Consistency of the corresponding adult survival

Adult survival rates were deduced from the recruitment rate estimates and the observed population growth rates, see Figure \@ref(fig:cor). In UK, values that are obtained from the counting method ranged from 0.21 [0.20; 0.22] to 0.68 [0.65; 0.72]. In France, survival rates vary from 0.39 [0.34; 0.44] to 1.08 [0.98; 1.17]. Like recruitment rates, there is significant year-to-year variability without clear temporal structure in both populations with a maximum amplitude of 0.69 in France (2.8-time variability). Unlike the recruitment rate, the survival rate in 2019 in France is not a low outlier.

As expected, survival rates estimated from UK sample data are consistently lower than estimates from the corresponding count data, Figure \@ref(fig:cor)A. The correlation between the two methods points out that the sampling method underestimates the survival rate by 32% compared to the counting method, Figure \@ref(fig:cor)B. Estimation precision is similar for the two methods. The sample time series shows that the adult survival is significantly lower after 2006, which corresponds to the start of the depletion of the UK population.

@@

########### Comparison to maximum recruitment rate

Maximum population growth rates are similar for the two populations, 1.45 [1.36; 1.55] and 1.47 [1.29; 1.68] for UK and France respectively, which corresponds to an increase of about 45% per year (Figure \@ref(fig:mgr)). After reaching a certain size (~ 1000 individuals), the UK population growth rate falls to 1.06 [1.04; 1.07], which corresponds to an increase of 6% per year, even though no harvest pressure was applied on it.

Upon the conservative assumption of a maximum adult survival uniformly distributed over 0.70 to 0.90, the maximum recruitment rate reaches 0.65 [0.51; 0.75] and 0.66 [0.45; 0.90] recruits per breeder for UK and France respectively. When there is no harvest, the estimated average recruitment rate of the French population is in the range of the proxy of the maximum recruitment rate, 0.51 [0.46; 0.57] recruits per breeder. When there is harvest, the average recruitment rates are 0.14 [0.12; 0.16] and 0.33 [0.31; 0.36] recruits per breeder for the UK and France respectively. Both values are significantly lower than the maximum recruitment rate, and it is noticeable that the average recruitment rate is lower for the UK population, which is in depletion over the count time series, than for the French population, which of the dynamic is mostly stable over the monitored period.

## Response of vital rates to two different harvest strategies

@@

dire que disantangle recruitment rate to survival bring capital information to understand the process of harvest on the population dynamics and brings knowledge sur les leviers pour mieux conserver une espèce 
on mieux la controler for invasive species

dire que grâce à ça, on peut fournir des variations interannuelles de survie et de recruitment, ce que ne permet pas toujours la cmr en raison des tailles d'échantillon.

Higher mortality on adult in GB -> better detection in winter;

Parler que le model peut etre amélioré en déclarant une proportion de mal à la repro dynamique qui prend en compte les années de fort recrutement vs les faibles recrutements.

Dire qu'il semble y avoir une meilleure detection en hiver au vu de la mortalité adulte entre UK & France

Count data in waterfowl that differentiate adult males from other individuals are successfully used in this study to infer estimations of recruitment rates. Since this parameter usually explains mostly the population growth in waterfowl [e.g. @Arnold2018], the non-invasive method introduced in this study might be considered as a candidate tool to monitor at low cost waterfowl populations. In the following section, one checks the accuracy of the method from indicators that were developed for this purpose. From the outputs on the two populations of ruddy duck, one discusses about the effects of the harvest on the recruitment rate and its implications on waterfowl management.

## Accuracy of the counting method

The recruitment rates from the counting method are in the same range for the two populations, which demonstrate a certain consistency in the results obtained from the method. The higher variability observed in France is not unexpected because the French time series is longer and covers a larger spectrum of harvest pressure than in UK. There is no temporal correlation between the two populations, so it suggests that the recruitment rate fluctuations are more likely influenced by regional-scale factors than large-scale ones. This is consistent with the literature because both nesting/hatching success and juvenile survival are conditional to the onset of lay, which is tightly related to local weather parameters, e.g. the spring temperature, the cumulative rainfall, and the water levels [@Folliot2017; @Blums2004; @Dzus1998].

Assuming that the sample data provide a good picture of the interannual variability of the recruitment rate, the good correlation with the outputs of the counting method demonstrates the ability of this method to track the recruitment rate variability. This result is quite robust because the correlation is obvious with only a five-year time series. The assumption of a constant adult sex ratio over time is then acceptable because a significant change in adult sex ratio among years would have biased the evolution of the recruitment rate. Since this ratio integrates many age-cohorts, the temporal autocorrelation is structurally strong. Significant variations can then occur only every multiple years, especially if the species is long-living. Consequently, it is not necessary to monitor and update the adult sex ratio on a yearly base. The adult sex ratio found here on a restricted time period is in line with the ruddy duck in its native area [0.62 in @Bellrose1980] and with other duck species [@Wood2021]. The robustness of the correlation between the two methods might be overrated because two successive population size values are parts of the recruitment rate, which mechanically scale the outputs. It explains why the correlation on the immature proportion, the other component of the recruitment rate, is a bit more noisy than for the recruitment rate. The uncertainty around the estimates is similar for the two methods even though the counting method is more indirect to infer on the recruitment rate, which is satisfying.

The counting method always provides recruitment rates significantly lower than the sampling method by a factor close to two, a result that is in line with the hypothesis that the harvest is generally selective towards immature individuals in waterfowl because they are more vulnerable than adults to hunters [@Bellrose1980; @Fox2014]. For hunted duck species in North America, the harvest-induced mortality on first-year immatures compared to adults is higher by a factor ranging from 1.3 to 2.6 [@Bellrose1980]. This is a robust result because the approach used to control ruddy ducks differ strongly from usual hunting practices and is apriori less selective. It is especially clear in the UK, for which control operations occurred in winter using long-range rifles that prevents the bird from escaping from the threat. The assumption on the accuracy of the counting method is then not excluded from the consistency of this result.

Annual adult survival rates deduced from observed population growth rates and recruitment rate estimates are on average in line with the literature on species of similar weight [@Nichols1997; @Krementz1997; @Buxton2004], which reflects that there is no apparent scaling issue. The assumption that male-like individuals corresponds only to adult males is then acceptable. This result supports that the counting method is poorly biased if it is. However, some adult survival rate values are outside the range of expected values even if they never significantly exceed the maximum possible value of one. A likely reason to these extreme values is a corruption of the closed population assumption on some years. Indeed, if there is an arrival from another ruddy duck population, the true recruitment cannot explain by itself the apparent population growth and indirectly leads to an overestimation of the adult survival.

Proxies of maximum recruitment rate are very close for the two populations. This strengthens that the ruddy duck species reached its intrinsic biological reproduction limit, at least given the environmental conditions of the western Palearctic. The consistency of this intermediate result suggests that the population size estimation is quite good for both populations and demonstrates that their evolution is well tracked even with a few individuals. The average recruitment rates estimated from the counting method, with and without harvest, are never higher to the proxies of maximum recruitment rate for both populations. The assumption on the accuracy of the counting method is then again not excluded from the consistency of this result.

Overall, even if the true values of recruitment rates are not known to test the accuracy of the counting method, a beam of arguments support that the method is not heavily biased. This consists in a big difference with the sampling method as demonstrated in @Fox2014, even when one considers a age structure picture that is limited to the end of the hunting season to limit the bias [@Fox2016].

## Influence of harvest on the recruitment rate

Only the French population was monitored over a period covering years without and with harvest. The harvest pressure, that is defined into a binary variable (no harvest / harvest), affects on average negatively both the recruitment rate and the adult survival for this population, as expected. If it explains only 4.6% of the variance of the adult survival, this value raises to 19.3% for the recruitment rate of the French population. The harvest pressure applied in France is thus a factor that likely induces a strong juvenile mortality that impacts the recruitment rate, but a negligible mortality on the adults. This conclusion is obviously opposed to the hypothesis that the heterogeneity in immature individuals would dampen the effect of the harvest pressure on the resulting recruitment rate. Two points can however explain this apparent discrepancy. Since the management targeted the eradication of this ruddy duck population, the harvest on immatures might be simply far higher than what the population could compensate and finally affects the recruitment rate. Also, the population started to be controlled when it was still growing at a high rate, so immature individuals were not suffering from density-dependent effects and were consequently sensitive to any additional mortality sources, like the new harvest pressure for instance. The low impact of the harvest on adult survival compared to the recruitment rate might be due to the timing of the management strategy adopted in France. Since the culling efficiency is poor on the wintering spot, most of control operations occurred in summer just after the reproduction, when the proportion of immature birds is the largest.

>>>>

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]

@@

  **Keywords**: fecundity - productivity - reproductive success - age ratio - vital rate - duck
output: 
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
    bookdown::word_document2:
      toc: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

We monitored the demography of Ruddy duck populations in both Great Britain and France thanks to exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition to these exhaustive counts, censuses that distinguished female-like from male-like individuals (Figure \@ref(fig:count)) were performed from 2006 to 2012 in Great Britain, which corresponded to a strong decline of the population. In France, the monitoring spread over 1999, 2001-2009, and 2012-2019. This population grew rapidly during the first years of the monitoring and then stabilized from 2006, which corresponds to a period with a strong control effort. During these years, the eradication effort, under control of the authorities in both countries, was recorded. Even if the age ratio in the hunting bag was uncertain before 2009 in France, we can qualify the harvest pressure on adults in three categories for both countries: "no harvest" before 1999, "low harvest" between 1999 and 2005, and "high harvest" from 2006 (see Figure \@ref(fig:sup) in Supplementary Materials).

@@

In waterfowl, the survival is commonly higher in adult males than in adult females. Adult sex ratio is then generally biased towards males (Figure \@ref(fig:pfem), @Wood2021). As a consequence, deducing the immature proportion within a population from the observed proportion of adult males is not straightforward because the adult proportion is not just twice as much as the observed proportion of adult males. There were no counts differentiating male-like from female-like during the breeding period to estimate properly adult sex ratio. But sex identification in adults from hunting bags was available for both Ruddy duck populations. However, the small population size in France prevented from getting precise adult sex ratios. A comparison over months in Great Britain showed no difference in adult male proportion, so data from adults collected over the whole year were used to estimate the male proportion. A comparison of male proportion among years with more than 500 samples did not show significant interannual differences. Even if the adult sex ratio may vary over a long time range, it is relatively stable over a few years [@Wood2021]. As a consequence, hunting bag data from all adults were pooled to estimate the male proportion in adults, see Equation \@ref(eq:pm).

@@

The maximum growth rate occurred for both populations during their expansion phase, when the evolution of the population size was the steepest. To estimate robust maximum population growth rates for both populations, we smoothed annual population growth rate over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L). To do so, we rejected Great Britain data before 1972 since the size estimate of this population is noisy below 50 individuals (Figure \@ref(fig:count)). After reaching the threshold of 1000 individuals, the Great Britain (GB) population growth observed a strong inflexion whereas no harvest pressure was applied. This observation suggests that a negative density-dependence process might occur beyond 1000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in Great Britain, i.e. 1972-1981. For the French (FR) population, the sequence without harvest spread over the period 1994-1999.

@@

Even if both vital rates vary in a likely range following the validation methodology previously described, it does not prove that the interannual variability is properly tracked. In order to validate this aspect, we compared the outputs of the count-based method, i.e. the immature proportion and both vital rates, to outputs based on data from hunting bags. Only data from years with more than 100 individuals harvested in winter were selected. It covers 5 years of the count time series of the GB population. The presence of the bursa of Fabricius enables immature to be identified [@Hochbaum1942]. The estimation of the immature proportion in hunting bags is then straightforward, see Equation \@ref(eq:pis). The adult survival and the recruitment rate were then estimated by combining this proportion and Equation \@ref(eq:vr). We discussed the ability of the count-based method to accurately track the interannual variability by comparing the outputs of the two methods.

@@

## Impact of harvest strategies on vital rates {#HARV}

Over the high harvest period, i.e. from 2006, control strategies differed from Great Britain to France. In Great Britain, harvest occurred mostly in winter, i.e. before the breeding period (53.4% of adults were shot before the 30^th^ May), whereas it was done mostly in summer in France, i.e. during and after the breeding period (81.6% of adults were shot after the 30^th^ May). We compared the population growth of the two populations in light of the difference between the two strategies and we assessed the respective response of the vital rates by comparing the average values over the high harvest period to proxies of maximum vital rates estimated when both populations reached their maximum growth (see Section \@ref(VAL)). In France, a LIFE project was implemented from early 2019 to intensify the harvest pressure, especially in winter and spring. We then excluded the FR time series from 2019 in order to provide a comparison of homogeneous harvest strategies.

For all the sub-models in Section \@ref(MM), the Bayesian framework was used for its efficiency and simplicity to propagate error through the parameters. We used uninformative priors on all parameters. As the maximum growth rate is an life history trait expected to be stable among populations of a same species, we used a uninformative hierarchical prior for this parameter. We generated three chains of length 500000, with a thinning of 10 to avoid autocorrelation in the samples, and we discarded the first 2000 samples as burn-in. Chain convergence was assessed using the Gelman and Rubin convergence diagnostic (R<1.1, @Gelman1992). We fit the models using NIMBLE [@Valpine2017] run from R [@RCT2022]. Data and code are available here: https://github.com/adri-tab/Ruddy_duck_vital_rates. The values **X[Y; Z]** reported in results are the medians and the associated boundaries of the 95% confidence interval of posterior distributions. The median was preferred to the mean because of its robustness to skewed distribution.

@@

## Estimates of vital rates

The method successfully provides estimates and associated uncertainty of immature proportion in both wintering populations. The adult survival rate and the recruitment rate, i.e. the two components of the population growth rate, are also successfully estimated by the method (Figure \@ref(fig:vital)).

@@

The male proportion in adults, which is a prerequisite to infer on the immature proportion, is estimated at 0.60 [0.59; 0.61]. The immature proportion varies from 0.16 [0.07; 0.24] to 0.54 [0.44; 0.62]. The lower values are similar for the two populations, but the range of the immature proportions in the GB population is far more limited (upper value : 0.30 [0.14; 0.45]). Conversely from the FR population, there is no significant difference from a year to another because all estimates overlap each other. We can notice that the immature proportion for the GB population is quite stable over the time series available, but a slight decrease is observed for the FR population.

The adult survival rate varies from 0.33 [0.29; 0.37] to 1.01 [0.89; 1.14]. All estimates are not significantly outside the range of a survival rate defined without immigration [0; 1]. There is no trend in adult survival for both populations, but we can notice two main differences. The GB population has the lowest values of adult survival rate, and even if there are significant difference among years for both populations, the variability for the GB population is a bit lower.

The recruitment rate varies from 0.09 [0.04; 0.13] to 0.75 [0.63; 0.87]. All estimates are above 0, meaning there is no senseless estimation. The maximum recruitment rate being 0.68 [0.36; 0.78] (see Section \@ref(MAX)), all estimates are not significantly outside the range of a recruitment rate defined without immigration [0; 0.78]. Similarly to the immature proportion, there is no trend for the GB population, but a decrease is observed for the FR population even if it is more noisy than for the immature proportion. Similarly to the survival rate, the GB population has low values of recruitment rate, and even if there are significant difference among years for both populations, the variability for the GB population is far lower than for the FR population.

The lower variability of both vital rates for the GB population explains its steady population growth over the considered time series compared to the FR one (Figure \@ref(fig:count)). In the FR population, the range of the recruitment rate (min/max difference: 0.66) is higher than the range of the survival rate (min/max difference: 0.58). Conversely, the range of the survival rate (min/max difference: 0.46) in the GB population is far higher than the range of the recruitment rate (min/max difference: 0.08). This reflects that the variability of the population growth rate for the FR population is mostly driven by the recruitment rate whereas the main source of variability is the adult survival for the GB population.

## Likelihood of year-to-year variability in vital rates

From the 5 years available in the GB population, immature proportions estimated from the count-based method is positively correlated to estimates from hunting-bag data (Figure \@ref(fig:cor)). The hypothesis that the proportion of female-like individuals in the wintering populations reflects the age structure of these populations is then supported by this result. This correlation is even stronger when considering vital rates (Figure \@ref(fig:cor)). This result is expected since the two methods estimating these vital rates have a component in common, the growth rate of the population (see Equation \@ref(eq:vr)).

It is noticeable that a 1:1 correlation is never obtained. Immature proportion is lower in counts than in hunting bags. This result is not surprising because immature individuals are expected to be more vulnerable to hunting than adults and might then be over represented in hunting bag data. As a consequence, the correlation coefficients for both vital rates differ also from 1. This difference between the two approaches does not question the previous validation of the ability of the count method to track the interannual variability of the vital rates.

@@

## Response of vital rates to two different harvest strategies {#MAX}

Under no harvest pressure, the maximum growth rates of the two populations are very similar (Figure \@ref(fig:mgr)), respectively 1.45 [1.36; 1.55] and 1.52 [1.25; 1.88] for the GB and the FR population. It corresponds to an increase of 45% per year and 52% per year respectively. Based on literature, we assumed a maximum survival rate similar for both populations 0.85 [0.7; 1]. We deduced a maximum recruitment rate of 0.60 [0.42; 0.78] for the GB population and of 0.68 [0.36; 1.06] for the FR population.

Even if a significant part of culled individuals were not aged for some years, the proxies of the harvest pressure show significant shifts over time (Figure \@ref(fig:sup)). This led to consider a high harvest pressure period from 2006 with no clear difference between the two populations because the signal is very noisy. Even if the harvest pressure seems comparable, the harvest strategies differ significantly between the two populations, see Section \@ref(HARV). Under high harvest pressure, the growth rate drops to 0.56 [0.53; 0.59] and 0.97 [0.94; 1.00] for the GB and the FR population respectively, which corresponds to a decrease of 44% per year and 3% per year. This result shows that the GB population significantly decreases during the high harvest pressure whereas the FR population is stabilized.

The adult survival rate under high harvest pressure is 0.44 [0.42; 0.46] and 0.62 [0.59; 0.64] for the GB and FR populations respectively, when the recruitment rate decreases to 0.13 [0.11; 0.15] and 0.32 [0.29; 0.34] respectively. The drop of the growth rate of the GB population corresponds to a similar average drop in adult survival rate (-0.41) and in recruitment rate (-0.47). The stabilisation of the FR population is more due to a decrease of the recruitment rate (-0.36) than from a drop of the adult survival (-0.23). For both populations, the response to the harvest pressure is significant, meaning that the compensation of the population to harvest mortality is limited if not null. Vital rates of the GB population are more affected than the FR population. Unsurprisingly, the FR harvest strategy focusing on the postbreeding period affects mostly the recruits. However, the prebreeding harvest strategy applied on the GB population that targets adults performs to deplete strongly the recruitment to a very low level (0.13 recruits produced per breeder).

@@

The delayed sexual dichromatism in a waterfowl species is used here to develop and test the validity of a new non-invasive method aiming to break down the apparent growth of a population into its two structural components: the adult survival rate and the recruitment rate. Disentangling the contribution of these two vital rates is a key to understand the response of a population to management measures. The development of the method holds on a unique dataset of two comparable populations of a same species, the Ruddy duck. Both populations were tracked from their introduction in a similar environment to a period of intense harvest pressure. This common history trajectory enables to evaluate the maximum growth of this species, but also its response to harvest pressure. As the harvest strategy differs between the two populations, we detail the similarities and the discrepancies of the responses of the vital rates, and open the discussion to the generalization of this method and its implications for waterfowl management.

## Accuracy of the method to estimate vital rates

The vital rate estimates of the two populations belongs to the same range, which demonstrate a certain consistency in the results obtained from the method. The higher variability observed in France is not unexpected because the FR time series covers a larger spectrum of harvest pressure than the GB time series. There is no temporal correlation between the two populations, suggesting that fluctuations of the vital rates are more likely influenced by regional factors than large-scale ones. This is consistent with the literature because both nesting/hatching success and juvenile survival are conditional to the onset of laying, which is tightly related to local weather parameters, e.g. spring temperature, cumulative rainfall, and water levels [@Folliot2017; @Blums2004; @Dzus1998].

The estimates of adult survival rates correspond to values observed in the literature on species of similar weight [@Nichols1997; @Krementz1997; @Buxton2004]. This reflects that there is no apparent scaling issue of this method based on count data. The assumption that male-like individuals correspond only to adult males is then acceptable. This result supports that the count-based method is poorly biased if it is. However, some adult survival rate values are outside the range of expected values even if they are never significantly over 1. A likely reason to these extreme values is a corruption of the closed population assumption on some years. Indeed, if there is an arrival from another Ruddy duck population, the true recruitment cannot explain by itself the population growth and leads to an adult survival combining true adult survival and adult immigration. For the FR population, upper outliers of recruitment rates and survival rates are observed in 2002 and 2012, suggesting immigration events on these years.

The theoretical maximum of the population growth rate is a potential which is determined by intrinsic life history traits of a species [@Dmitriew2010]. This potential is expected to be similar to all Ruddy duck populations anywhere in the world. Realised maximum growth rates that are observed in this study are very close for the two populations. It suggests that the Ruddy duck species reached its intrinsic biological reproduction limit, at least given the environmental conditions of the western Palearctic, i.e. about 0.6 recruits per breeder. The consistency of this intermediate result suggests that the population size estimation is quite accurate for both populations and demonstrates that their changes are well tracked even with a few individuals. The average recruitment rates estimated from the count-based method, with and without harvest, are never higher to the proxies of maximum recruitment rate for both populations. The assumption on the accuracy of the count-based method is then again not excluded from the consistency of this result.

Assuming that hunting-bag data provide a good picture of the year-to-year variability of age structure of a population, the strong correlation with the outputs of our method demonstrates its ability to track interannual fluctuations. This result is robust because such correlation is obvious despite being based on a short time series. This strong relationship is also satisfying because it demonstrates that the assumption of a constant adult sex ratio is not corrupted. The temporal autocorrelation of the adult sex ratio might be explained because it integrates many age-cohorts, which makes it structurally strong. Even if significant variations of the adult sex ratio can still occur over long time periods, this result demonstrates that it is not necessary to monitor and update it on a yearly base. The adult sex ratio found here on a restricted time period is in line with the Ruddy duck in its native area [0.62 in @Bellrose1980] and with other duck species [@Wood2021]. The count-based method always provides recruitment rates significantly lower than the method based on hunting-bag data by a factor close to two, a result that is in line with the hypothesis that the harvest is generally selective towards immature individuals in waterfowl because they are more vulnerable than adults to hunters [@Bellrose1980; @Fox2014]. The gap of vulnerability among ages observed here is realistic because the harvest-induced mortality on first-year immatures compared to adults is higher by a factor ranging from 1.3 to 2.6 for hunted duck species in North America [@Bellrose1980].

Overall, even if the true values of vital rates are not known to test properly the accuracy of the count-based method, a beam of arguments supports that the method is not heavily biased. This consists in a big difference with the method based on hunting-bag data as demonstrated in @Fox2014, even when one considers an age structure picture that is limited to the end of the hunting season to limit the bias [@Fox2016].

## Population response to managment measures

Both populations respond to the harvest pressure by a large drop of their growth, meaning that the the harvest effort is efficient to control this species. The response of the GB population differs from the FR one because it displays a strong decrease when the FR population is stabilized. The analysis of the growth alone does not provide clues to understand the mechanisms explaining the difference of the responses between the two populations. Indeed, the bigger drop of growth of the GB population might be either explained either by a decrease of the adult survival or of the reproduction success, or even of both components.

The vital rates that are estimated thanks to the count-based method developed in here demonstrates that the harvest pressure affects on average negatively both the adult survival and the recruitment rate for the two populations. The harvest pressure occurring mostly before the breeding period for the GB population targets by definition future breeders, and consequently induces mortality on these individuals and prevent them from reproducing. We observe that adult survival and recruitment rate decrease equally, which is a result in line with the expectation of this harvest strategy. For the FR population, the harvest pressure occurs mostly during the postbreeding period. This strategy should induce a higher mortality on immatures as they are expected to be more vulnerable to hunting [@Bellrose1980; @Fox2014]. However, this anthropogenic mortality on immatures might be compensated by a higher survival of the future recruits because the competition for resources decreased [@Cooch2014]. The recruitment rate of the FR population being more affected by 50% than the adult survival, the results support the hypothesis that no or only a limited compensation occurs. This conclusion is even strengthened by the result on the GB population. Indeed, we should expect a decrease of the recruitment rate lower than the adult survival rate if immatures would take advantage of additional resources that unexploited by culled adult individuals. The equal decrease of the two vital rates supports then that there is no compensation    at all. This might be explained because the population size is far lower than the carrying capacity of the available habitats, so there is no competition for resources that potentially enhances the survival of the remaining immature individuals [@Peron2013].
